A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-18 podc_paper/podc-submission.tex
working on related work
got rid of sample .bib entries
moved algorithm description into separate file algorithm.tex

\newcommand{\here}[1]{\bf{[[#1]]}}

@@

\input{introduction}
\input{related}

\section{Queue Implementation} \label{DescriptQ}

In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently. We design a linearizable wait-free queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of linearization. We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.

There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes. The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree}). Each process has a leaf in which the operations invoked by the process are stored in order. When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root. Each node of the tree keeps an ordering of operations propagated up to it. All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering. 
\begin{figure}[hbtp]
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption[Tournament tree.]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored. Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}  
\end{figure}

To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$. We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}). A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to $n$. We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
We sketch the argument here.

\begin{figure}[hbtp]
\begin{center}
\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{Before the \nf{Refresh}.}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{After the \nf{Refresh}.}
\end{subfigure}
\hspace{6em}
\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append. Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
\end{center}
\end{figure}

\begin{figure}[hbpt]
  \center\includegraphics[width=4in]{pics/doublyrefresh-drawio.png}
  \caption[Two consecutive failed \nf{Refresh}es by a process.]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
\end{figure}

We use \nf{CAS} (Compare\&Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}). This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}. Process $K$'s \nf{Refresh} helps to append the new operations that were in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended. After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node. So, with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization. This cooperative solution allows us to overcome the CAS Retry Problem.

It is not efficient to explicitly store the sequence of operations in each node because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
Instead we use an implicit representation of the operations propagated together. Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root. It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}). Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions on fixed-size objects in each \tt{Refresh}. To do that, we introduce \it{block}s. A block stores information about the operations propagated by a \nf{Refresh}. It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure. See Figure~\ref{fig:block} for an example. A node stores an array of blocks of operations propagated up to it. A propagate step  aggregates the new blocks in the children into a new block and stores the new block in the parent. We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.  In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently. Thus, there are at most $p$ operations in a block. Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way, if we know the number of operations from the left child and the number of operations from the right child in a block, we have a complete ordering of the operations.

\begin{figure}[hbtp]
\begin{center}
\begin{tikzpicture} [level 1/.style={level distance=1.8cm,sibling distance=0.6cm}, level 2/.style={level distance=1.4cm,sibling distance=0.4cm}]

\Tree [.{$\big\{op_2^1,op_1^1,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_2^2, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_1^1\big\},\big\{op_2^2\big\},\big\{op_1^2\big\}...$ }
      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ... \\ \hline \end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ... \\ \hline\end{tabular}} ] ]
\end{tikzpicture}
\caption[The operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right. In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}. For example $op_4^1$ and $op_3^2$ have propagated  together in one \nf{Propagate} step and they will be propagated up to the root together. Since their execution time intervals overlap, they can be linearized in any order.}
\end{center}
\end{figure}

\begin{figure}[hbtp]
\begin{center}
\begin{tikzpicture}
[level 1/.style={level distance=3.3cm,sibling distance=0cm},
	level 2/.style={level distance=2.5cm,sibling distance=0cm},
	level 3/.style={level distance=1.7cm,sibling distance=2cm}]
\begin{small}

\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
 [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
 [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
            [.{\begin{tabular}{||c|c||c||c||}  \hline (2,1) & (5,0) & (4,2) & (9,7) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
\end{small}
\end{tikzpicture}
\end{center}
\caption[The operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations. Blocks between two lines $||$ are propagated together to the parent. Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively. For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child. The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child. (5,3) is the superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
\end{figure}

So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root. With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$. However, this approach is not enough for an efficient queue.
We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization. We explain how to implement (1) and (2) in poly-logarithmic steps.

After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in the descendent of the root that contains that $i$th operation. When we reach a block in a leaf, the operation is explicitly stored there. To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block. We also store pointers to determine the range of subblocks of a block to make the binary search faster. In each block, we store the prefix sum of operations from the left child and from the right child. Moreover, for each block, we store two attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of the last left and right subblock (see Figure \ref{fig::pointer}). We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and \nf{end\sub{right}} indices of a block and its previous block reduce the search range size to \textsc{O}$(p)$.

To compute the rank in the root of an operation in a leaf, we need to find the superblock of the block that the operation is in. After a block is installed in a node we store the approximate index of its superblock in it to make this faster.

\begin{figure}[hbtp]
\centering
  \includegraphics[width=4in, height=1.4in]{pics/pointers}
  \caption{Each block stores the index of its last subblock in each child. \label{fig::pointer}}
\end{figure}

In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue. In the general case a dequeue returns \nf{null} if and only if the queue size after the previous operation is 0. We refer to such a dequeue as a \it{null dequeue}. If the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue. Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues before the operation. So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.

In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific position.  We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations. Consider the following operations, where operations in a cell are concurrent.
\begin{table}[hbtp]
\centering
\begin{small}
\begin{tabular}{c|c|c|c}
    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
  \end{tabular}
\end{small}
\caption{An execution on a queue separated by the operations in the root blocks.}
\end{table}

\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
Now, we claim that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}. We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.

\begin{table}[hbtp]
\centering
\begin{small}
  \begin{tabular}{c|c|c|c|c}
    \hline &\texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
    \#\tt{Enq}s & 0 & 3 & 1 & 1 \\ \hline
    \#\tt{Deq}s & 1 & 1 & 1 & 4 \\ \hline
    Size at end & 0 & 2 & 2 & 0 \\ \hline
  \end{tabular}
\end{small}
  \caption{Augmented history of operation blocks on the queue.}
\end{table}

\noindent The size of the queue after the $b$th block in the root could be computed as
$$\textrm{max}\Big(\textrm{size after }(b-1)\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$
\noindent Moreover, the total number of non-null dequeues in blocks $1,2,...,b$ in the root is
$$ \sum_{i=1}^{b} \#\textrm{\tt{Enqueue}s in }i\textrm{th block} - \textrm{size after }b\textrm{th block}.$$

Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is
\begin{center}
    \#\textrm{non-null } \nf{Dequeue}s in  blocks $1,2,...,b-1 +  $index of the \nf{Dequeue} among B\textrm{'s} \nf{Dequeue}
\end{center}
if \vspace{-3em}
\begin{center}
    $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} $\\$- \textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
\end{center}
Otherwise, the response would be \nf{null}.

\subsection{Details of the Implementation}
Section \ref{algQ} gives the pseudocode for the queue implementation. It uses the following two types of objects.
\paragraph{\tt{Node}} 
 In each \texttt{Node} we store pointers to its parent and its left and right child, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.

\paragraph{\tt{Block}}
 The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf. If it is in a leaf, we use a \nf{LeafBlock} which stores one operation. If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$. The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the left subblocks of the block prior to $B$ ended. The right subblocks of $B$ are defined similarly. In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}. The \texttt{sum\textsubscript{enq-left}} field is the total number of \nf{Enqueue} operations in the blocks before the last subblock of $B$ in the left child. The other fields' semantics are similar. The \nf{end\sub{left}} and \nf{end\sub{right}} field store the index of the last subblock of a block in the left and the right child, respectively. The approximate index of the superblock of non-root blocks is stored in their \nf{super} field. The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.

We now describe the routines used in the implementation.

\paragraph{\tt{Enqueue($e$)}}
An \nf{Enqueue} operation does not return a response, so it is sufficient to propagate the \nf{Enqueue} operation to the root and then use its position in the linearization for future \nf{Dequeue} operations. \nf{Enqueue($e$)} creates a \nf{LeafBlock} with $\nf{element}=e$, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and then appends it to the tree.

\paragraph{\tt{Dequeue()}}
\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and appends it to the tree. Then, it computes the position of the appended \nf{Dequeue} operation in the root using \nf{IndexDequeue} and after that finds the response of the \nf{Dequeue} by calling \nf{FindResponse}.

\paragraph{\tt{Append($B$)}}
The \nf{head} field is the index of the first empty slot in \nf{blocks} in a \nf{LeafBlock}. There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it. \nf{Append($B$)} adds $B$ to the end of the \nf{blocks} field in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}. When \nf{Propagate} terminates, it is guaranteed that the appended block is a subblock of a block in the \nf{root}.

\paragraph{\tt{Propagate()}}
\nf{Propagate} on node $n$ uses the double refresh idea described earlier and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}. Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively until it reaches the root.

\paragraph{\tt{Refresh()} and \tt{Advance()}}
The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}. The variable \nf{h} is read from $n$\nf{.head} at Line \ref{readHead}. The new block created by \nf{Refresh} will be inserted into $n\nf{.blocks[h]}$. Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children. \nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recently appended blocks. The reason behind this helping is explained later when we discuss \nf{IndexDequeue}. After helping to \nf{Advance} the children, a new block called \nf{new} is created in Line \ref{invokeCreateBlock}. Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate, and it is unnecessary to add an empty block to the tree. Later we will use the fact that all blocks contain at least one operation. Line \ref{cas} tries to install \nf{new}. If it was successful, all is good. If not, it means someone else has already put a block in $n$\nf{.blocks[h]}. In this case, \nf{Refresh} helps advance $n$\nf{.head} to \nf{h+1} and update the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.

\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
The block \nf{new} is created in Line \ref{initNewBlock} and its fields are filled similarly for both left and right directions. The variable  \nf{index\sub{prev}} is the index of the block preceding the first subblock in the child in direction \nf{dir} that is aggregated into \nf{new}. Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in the child. Then \nf{sum\sub{enq-dir}} is computed from the sum of the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}. The field \nf{sum\sub{deq-dir}} is computed similarly. Then, if \nf{new} is going to be installed in the \nf{root}, the \nf{size} field is also computed.

\paragraph{\tt{IndexDequeue($b,i$)}}

A call to $n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block in the root of the $i$th \nf{Dequeue} of the $b$th block of $n$. Let $R_n$ be the successful  \nf{Refresh} on node $n$ that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}. Let $par$ be $n$\nf{.parent}. Without loss of generality, assume for the rest of this section that ${n}$ is the left child of $par$. Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ as a subblock of its created block in Line \ref{invokeCreateBlock}. Let $j$ be the index of the block that $R_{par}$ puts in $par$\nf{.blocks}.

Since the index of the superblock of ${B}$ is not known until $B$ is propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while creating $B$. One approach for $R_{par}$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $par$. This solution would not be efficient because there might be up to $p$ subblocks that $R_{par}$ propagated, which need to update their \nf{super} field. However, intuitively, once $B$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation. If we know the approximate position of the superblock of $B$ then we can search for the real superblock when it is needed. Thus, \nf{$B$.super} does not have to be the exact location of  the superblock of $B$, but we want it to be close to $j$. We can set \nf{$B$.super} to \nf{$par$.head} while creating $B$, but the problem is that there might be many \nf{Refresh}es on $par$ that could happen after $R_n$ reads $par$\nf{.head} and before propagating $B$ to $par$. If $R_n$ sets \nf{$B$.super} to \nf{$par$.head} after appending $B$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $B$ and before setting \nf{$B$.super}. In this case, the next \nf{Refresh}es on $n$ and $par$ help fill in the value of \nf{$B$.super}.

Block $B$ is appended to \nf{$n$.blocks[$b$]} on Line \ref{cas}. After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}. We shall show that this is sufficient to ensure that \nf{$B$.super} differs from the index of $B$'s superblock by at most~1.

\paragraph{\tt{FindResponse($b, i$)}}
To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not. If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}. If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is the response to the \nf{Dequeue}. Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block), we find the block and position containing the \nf{Enqueue} operation using \nf{DoublingSearch} and after that \nf{GetEnqueue} finds its \nf{element}.

\paragraph{\tt{GetEnqueue($b,i$)} and \tt{DoublingSearch($e, end$)}}
We can describe an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation within that block. If we know the block and rank within the block of an operation, we can find the subblock containing the operation and the operation's rank within that subblock in poly-log time. To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
We also know the \nf{e}th \nf{Enqueue} is in \nf{root.blocks[1..end]}. \nf{DoublingSearch} uses doubling to find the range that contains the answer block (Lines \ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the required indices with a binary search (Line \ref{dsearchBinarySearch}).
A call to \nf{$n$.GetEnqueue($b,i$)} returns the \nf{element} of the $i$th enqueue in the $b$th block of $n$. The range of subblocks of a block is determined using the \nf{end\sub{left}} and \nf{end\sub{right}} fields of the block and its previous block. Then, the subblock is found using binary search on the \nf{sum\sub{enq}} field (Lines \ref{leftChildGet} and \ref{rightChildGet}).

@@

@@

\bibliographystyle{plain}
\bibliography{sample-base.bib}

>>>>

\newcommand{\here}[1]{\bf{[[#1]]}}
\newcommand{\ceil}[1]{\lceil #1 \rceil}

@@

\input{introduction}
\input{related}
\input{algorithm}

@@

Tarjan \cite[Sec.~4.2]{Tar83} described a split algorithm for red-black trees that runs in logarithmic time.

@@

\bibliographystyle{plain}
\bibliography{queues.bib}

### 2022-12-18 podc_paper/related.tex
moved algorithm description into separate file algorithm.tex

\paragraph{Fetch\&Increment Objects}
Progress has been made on obtaining sublinear-time implementations of other objects 
with consensus number 2.
Two papers \cite{ERW12,10.1007/978-3-642-41527-2_20}
implemented fetch\&increment objects from LL/SC objects
using a polylogarithmic number of steps per operation.
Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
which uses a tournament tree to keep track of the fetch\&increment operations
that have been performed.  However, our construction requires much more intricate
data structures, since the state of a queue cannot easily be represented as succinctly
as the single-word state of a fetch\&increment object.

>>>>

\paragraph{Fetch\&Increment Objects}
Progress has been made on obtaining sublinear-time implementations of other objects 
with consensus number 2.
Two papers \cite{ERW12,10.1007/978-3-642-41527-2_20}
implemented fetch\&increment objects from LL/SC objects
using a polylogarithmic number of steps per operation.
Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
which uses a tournament tree to keep track of the fetch\&increment operations
that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.
However, our construction requires more intricate
data structures to represent sets of operations, since the state of a queue cannot easily be represented as succinctly
as the single-word state of a fetch\&increment object.
